Optimizing Abstract$^{\text 2}$ Machines

### 2012-11-02 paper/paper.tex
Elaborated on compound data abstraction.

TODO: {\tt (cons A-addr D-addr)}, {\tt qpair}, {\tt opaque-atom}.

>>>>

\subsubsection{Na\"ively}

The uniform way they approach a simple abstraction strategy is to cut
recursion out of the data definition by tying the recursive knot
through the abstract store. For Scheme, the grammar for values looks like the following:

\begin{alltt}
Value ::= #t | #f | (cons Value Value) | '() | ...
\end{alltt}

Upon evaluating a {\tt cons} application, we instead allocate two
addresses $a$ and $d$, join them to the respective values in the
store, and return the flattened {\tt (cons $a$ $d$)} value. Since
these addresses are all distinguished at different syntactic
callsites, and quoted lists really are sugar for a sequence of calls
to {\tt cons}, this abstraction explodes the value space. Analyzing a
function that counts the number of atoms in a literal s-expression
would actually interpret that function at least that number of times
(more because of intermediate conses). Indeed, even in our fastest
parameterization, this abstraction causes the analysis of Boyer to be
430 times slower than the approach we will now describe.

\subsubsection{Less na\"ively and less precisely}

The number of syntactic uses of {\tt cons} versus implicit uses via
literal lists is dramatically smaller in typical Scheme programs. We
use the above abstraction for these syntactic uses, but choose a
different strategy for quoted lists. Quotation is special because it
cannot introduce function values or mutable values, which is important
to enhancing our technique's soundness to conceptual complexity ratio.

There are a few steps to consider:
\begin{itemize}
 \item{Define special value lattice elements for compound data domains that can be quoted
       (e.g. {\tt Pair} for $\left\lbrace ({\tt cons} a_i d_i)\right\rbrace_i$, {\tt IVector} for immutable vectors, etc.)}
 \item{Define a ``larger'' value lattice element for all quotable data, {\tt QData}}
 \item{Interpret {\tt (quote (a ...))} as {\tt (qlist a ...)}, a new primitive function defined in figure \ref{fig:qlist},
       and similar definitions for immutable vectors.}
 \item{Extend the $\Delta$ axioms to include conservative meaning for these new values (e.g. {\tt (car QData)} = {\tt QData}, {\tt (add1 QData)} = {\tt Number} and log ``possible type error'') and allow them to allocate addresses and change the store}
\end{itemize}

\begin{figure}
\begin{align*}
\Delta(\msto, qlist) &= (\mbox{{\tt '()}}, \msto) \\
\Delta(\msto, qlist, v ..._+) &= (({\tt cons}\ a\ d), \msto') \\
\qquad \mbox{where } \msto' &= \msto\sqcup[a \mapsto \bigsqcup(v ...)] \\
                            &\qquad \sqcup[d \mapsto \lbrace \mbox{{\tt '()}}, ({\tt cons}\ a\ d)\rbrace] \\
\bigsqcup(v) &= v \\
\bigsqcup(v, vs ...) &= merge(v, \bigsqcup(vs ...)) \\
merge(v, v) &= v \\
merge(n, m) &= {\tt Number} \\
merge(n, v) &= {\tt QData} \\
merge(({\tt cons}\ a\ d), ({\tt cons}\ a'\ d')) &= {\tt Pair} \\
merge({\tt Pair}, ({\tt cons}\ a\ d)) &= {\tt Pair} \\
merge({\tt Pair}, v) &= {\tt QData} \\
\vdots
\end{align*}
\caption{Quoted list primitive}
\label{fig:qlist}
\end{figure}

The alternative to {\tt QData} is to go higher in the lattice to
$\top$, which has more complicated meaning (Shivers' escape semantics) and
is overly approximate.
